\chapter{INTRODUCTION}

\section{Scope of this report}
This report outlines the development process of the programming language
implemented in accordance with the specification of AT70.07 Programming
Languages and Compilers --- Assignment~2. The project was divided into three
main components: the lexer and symbol table, the parser, and the type system
with an interpreter. These were structured to be developed consecutively rather
than simultaneously. By organising the work in sequence, each team member could
focus on an assigned component independently before passing it to the next
stage, which supported steady progress while reducing the need for overlapping
work against other academic commitments.

\section{Background and Motivation}
Programming languages provide the formal structure through which algorithms are
expressed and executed. A compiler or language processor is responsible for
turning source code into a form that can be analysed or executed while also
detecting errors as early as possible. Core compiler concepts such as lexical
analysis, parsing, abstract syntax trees, symbol tables, static type checking,
and interpretation are therefore fundamental topics in programming language
education.

The motivation for this project is to apply those concepts in a compact but
complete language implementation. By building a small statically-typed language,
the project demonstrates how syntax, semantics, and execution can be connected
in a coherent system. The assignment also provides an opportunity to study how
type inference can improve usability while still preserving compile-time error
checking.

\section{Language Overview}
The implemented language supports four primitive types: Integer, Float,
Boolean, and String. It uses static typing with inference, so programmers do
not need to write explicit type declarations. Expressions include arithmetic
operations with standard precedence and Boolean comparisons through equality and
inequality. The statement-level constructs include assignment, if-then-else,
while-loop, function definition, function call, return, and the built-in
\texttt{print()} function.

Programs are processed through a lexer, a recursive-descent parser, a type
checker, and an interpreter. The final system can be executed from the command
line or through a desktop UI that allows the user to enter a script, run it,
and inspect the resulting tokens, AST, inferred types, and output.

\section{Report structure}
The remainder of this report is organised as follows. Chapter~2 presents the
language design, token specification, grammar, and operator precedence rules.
Chapter~3 explains the type inference algorithm and the semantic rules used by
the checker. Chapter~4 presents system architecture and implementation: the end-to-end
pipeline (with a figure of the stages), project layout, technology stack, and
how the type checker, interpreter, and user interfaces attach to the shared
\texttt{pipeline.py} module. Chapters~5 and~6 present the front-end
implementation in pipeline order: lexical analysis and the symbol table, then
recursive-descent parsing and the AST. Chapter~7 discusses testing and
verification, including the automated
regression tests. Chapter~8 concludes the report and summarises the final
outcomes of the project.
