\chapter{LANGUAGE DESIGN AND GRAMMAR}

\section{Types}
The language supports four primitive types. These types are sufficient to cover
the assignment requirements while keeping the semantics simple and predictable.

\subsection{Integer}
The Integer type represents whole numbers such as \texttt{0}, \texttt{7}, and
\texttt{42}. Integer literals are written as one or more decimal digits without
quotation marks or a decimal point.

\subsection{Float}
The Float type represents real-valued numbers written with a decimal point, such
as \texttt{3.14} or \texttt{20.5}. Float values are used in arithmetic
expressions and may also arise through numeric promotion.

\subsection{Boolean}
The Boolean type represents logical truth values. The two Boolean literals are
\texttt{true} and \texttt{false}. Boolean values are used in conditions and as
results of comparison expressions.

\subsection{String}
The String type represents sequences of characters enclosed in double quotes,
for example \texttt{"hello"}. Strings can be assigned to variables and printed
through the built-in \texttt{print()} function.

\section{Static Typing}
The language uses static typing, which means type errors are detected before
execution rather than during program execution. No explicit variable type
declarations are required. Instead, the type checker infers types from literals,
assignments, operations, function calls, and return statements. Once a variable
type is established, later assignments must remain consistent with that type.

\section{Token Specification}
Table \ref{tab:token-specification} summarises the major token classes used by
the language.

\begin{table}[H]
\centering
\caption{Token Specification}
\label{tab:token-specification}
\begin{tabular}{ll}
\toprule
Token Category & Example Lexemes \\
\midrule
\texttt{INT\_LITERAL} & \texttt{0}, \texttt{10}, \texttt{42} \\
\texttt{FLOAT\_LITERAL} & \texttt{3.14}, \texttt{20.5} \\
\texttt{BOOL\_LITERAL} & \texttt{true}, \texttt{false} \\
\texttt{STRING\_LITERAL} & \texttt{"hello"} \\
\texttt{IDENTIFIER} & variable and function names such as \texttt{x}, \texttt{add} \\
Arithmetic operators & \texttt{+}, \texttt{-}, \texttt{*}, \texttt{/} \\
Comparison operators & \texttt{==}, \texttt{!=} \\
Assignment operator & \texttt{=} \\
Keywords & \texttt{if}, \texttt{else}, \texttt{while}, \texttt{def}, \texttt{return}, \texttt{print} \\
Delimiters & \texttt{( ) \{ \} , ;} \\
\bottomrule
\end{tabular}
\end{table}

\section{Context-Free Grammar}
The language grammar can be described by the context-free grammar
$G = (V_N, V_T, P, S)$, where $V_N$ is the set of non-terminals, $V_T$ is the
set of terminals, $P$ is the set of production rules, and $S$ is the start
symbol \texttt{program}. The grammar is implemented through recursive-descent
parsing functions.

\subsection{Program Structure}
\begin{lstlisting}
program   -> statement* EOF
statement -> assignment ';'
	   | if_stmt
	   | while_stmt
	   | func_def
	   | print_stmt ';'
	   | return_stmt ';'
	   | expr ';'
\end{lstlisting}

\subsection{Arithmetic Expressions}
\begin{lstlisting}
expr   -> term (('+' | '-') term)*
term   -> factor (('*' | '/') factor)*
factor -> INT_LITERAL
	| FLOAT_LITERAL
	| STRING_LITERAL
	| BOOL_LITERAL
	| IDENTIFIER
	| IDENTIFIER '(' args? ')'
	| '(' expr ')'
\end{lstlisting}

\subsection{Boolean Expressions}
\begin{lstlisting}
bool_expr -> expr '==' expr
	   | expr '!=' expr
	   | BOOL_LITERAL
\end{lstlisting}

\subsection{Statements}
\begin{lstlisting}
assignment  -> IDENTIFIER '=' expr
if_stmt     -> 'if' '(' bool_expr ')' block ('else' block)?
while_stmt  -> 'while' '(' bool_expr ')' block
print_stmt  -> 'print' '(' expr ')'
return_stmt -> 'return' expr
block       -> '{' statement* '}'
\end{lstlisting}

\subsection{Function Definitions and Calls}
\begin{lstlisting}
func_def -> 'def' IDENTIFIER '(' params? ')' block
params   -> IDENTIFIER (',' IDENTIFIER)*
args     -> expr (',' expr)*
\end{lstlisting}

Function calls use value parameter passing. This means the argument values are
evaluated before the call and copied into the function's local environment.

\section{Operator Precedence and Associativity}
Operator precedence is encoded structurally in the recursive-descent parser.
Comparison is lowest, addition and subtraction are next, and multiplication and
division have the highest precedence among binary operators. Arithmetic
operators are left-associative.

\begin{table}[H]
\centering
\caption{Operator Precedence and Associativity}
\begin{tabular}{lll}
\toprule
Level & Operators & Associativity \\
\midrule
1 & \texttt{==}, \texttt{!=} & comparison level \\
2 & \texttt{+}, \texttt{-} & left-associative \\
3 & \texttt{*}, \texttt{/} & left-associative \\
\bottomrule
\end{tabular}
\end{table}
