\chapter{TYPE INFERENCE ALGORITHM}

\section{Overview}
Type inference in this project means that the programmer does not write explicit
type declarations for variables or functions. Instead, the compiler infers
types from assignments, literals, operations, function calls, and return
statements. This keeps the language simple while still preserving static type
safety.

The type checker runs after parsing. It receives the AST and validates the
program before execution. Any inconsistent use of types is reported as a type
error, and execution is stopped.

\section{Type Environment}
The type environment is represented by the symbol table. Conceptually, it is a
mapping

\begin{center}
$\Gamma : \textit{Identifier} \rightarrow \textit{Type}$
\end{center}

The environment grows as the checker visits the AST. On first assignment, a
variable name is inserted with its inferred type. When a new block or a
function call is entered, a child environment is created so that nested scopes
can be checked independently while still accessing outer bindings when needed.

\section{Inference Rules}

\subsection{Literals}
Literal nodes have direct types:

\begin{align*}
\Gamma \vdash 42 &: \texttt{Integer} \\
\Gamma \vdash 3.14 &: \texttt{Float} \\
\Gamma \vdash \texttt{true} &: \texttt{Boolean} \\
\Gamma \vdash \texttt{``hi''} &: \texttt{String}
\end{align*}

\subsection{Arithmetic Expressions}
The arithmetic operators \texttt{+}, \texttt{-}, \texttt{*}, and \texttt{/}
accept only numeric operands. If both operands are \texttt{Integer}, the result
is \texttt{Integer}, except for division, which produces \texttt{Float}. If at
least one operand is \texttt{Float}, the result is \texttt{Float}. Any
non-numeric operand causes a type error.

\subsection{Boolean Expressions}
The comparison operators \texttt{==} and \texttt{!=} always produce
\texttt{Boolean}. Numeric comparison is allowed for arithmetic expressions. The
current implementation also permits Boolean-to-Boolean comparison because the
sample programs use \texttt{flag!=false}. Other combinations are rejected.

\subsection{Assignment Statement}
Assignments define variable types on first use. If a variable does not yet
exist in the type environment, the checker infers the type of the right-hand
side and inserts the variable into the environment. If the variable already
exists, the new value must have the same type.

\begin{center}
If $x \notin \Gamma$ and $\Gamma \vdash e : T$, then $\Gamma, x:T \vdash x = e$
\end{center}

\begin{center}
If $x:T \in \Gamma$ and $\Gamma \vdash e : T$, then $\Gamma \vdash x = e$
\end{center}

\subsection{If-Then-Else}
The condition of an \texttt{if} statement must be \texttt{Boolean}. The
then-block and else-block are checked in child scopes so that local operations
can be validated cleanly.

\subsection{While-Loop}
The condition of a \texttt{while} loop must also be \texttt{Boolean}. The loop
body is checked in a child scope and must not violate the types of variables
already introduced in the surrounding environment.

\subsection{Function Definition and Call}
Function parameter types are inferred from the first call to the function.
After that first call, subsequent calls must use arguments of the same types.
The return type is inferred from the \texttt{return} statements in the function
body. If the function returns inconsistent types, the checker reports an error.

Recursive calls are intentionally rejected in this implementation because they
were not required by the assignment and would require additional inference and
control-flow analysis.

\section{Type Error Reporting}
Type errors are reported with source positions so that the user can locate the
problem quickly. The format used by the system is:

\begin{center}
\texttt{[line X, col Y] TypeError: message}
\end{center}

Examples include assigning a \texttt{String} to an \texttt{Integer} variable,
using a non-Boolean condition in an \texttt{if} statement, or supplying the
wrong argument type to a function.

\section{Example Type Inference Traces}
Consider the following program fragment:

\begin{lstlisting}
x = 10;
y = 20.5;
z = x + y;
print(z);
\end{lstlisting}

The checker infers:
\begin{itemize}
	\item \texttt{x: Integer}
	\item \texttt{y: Float}
	\item \texttt{x + y: Float}
	\item \texttt{z: Float}
\end{itemize}

For a function example:

\begin{lstlisting}
def add(a, b) {
		return a + b;
}

result = add(2, 3.5);
\end{lstlisting}

The first function call infers \texttt{a: Integer} and \texttt{b: Float}. The
expression \texttt{a + b} is therefore \texttt{Float}, so the function return
type is inferred as \texttt{Float}. The variable \texttt{result} is also
inferred as \texttt{Float}.
