\chapter{SYSTEM ARCHITECTURE AND IMPLEMENTATION}

\section{Compiler Pipeline}
The implemented system follows a four-stage pipeline. Source code is first
converted into tokens by the lexer. The parser then constructs an AST from the
token stream. After parsing, the type checker validates and infers types across
the AST. Finally, the interpreter executes the checked program and emits output
line by line.

The full processing pipeline is:

\begin{center}
\texttt{Source Code $\rightarrow$ Tokens $\rightarrow$ AST $\rightarrow$ Type Check $\rightarrow$ Execution Output}
\end{center}

This design keeps syntax handling independent from semantic checking and
execution, which simplifies testing and debugging.

\section{Project Structure}
The main source files and their responsibilities are summarised below.

\begin{table}[H]
\centering
\caption{Main Source Files}
\begin{tabular}{ll}
\toprule
File & Responsibility \\
\midrule
\texttt{main.py} & Command-line runner for sample code or a script file \\
\texttt{ui.py} & Desktop UI for editing and running a script \\
\texttt{components/tokens.py} & Token classes and token categories \\
\texttt{components/lexica.py} & Lexical analysis \\
\texttt{components/ast\_nodes.py} & AST node definitions \\
\texttt{components/parser.py} & Recursive-descent parser \\
\texttt{components/ast\_printer.py} & AST renderer for debugging/reporting \\
\texttt{components/type\_checker.py} & Static type checker and inference \\
\texttt{components/interpreter.py} & Tree-walking interpreter \\
\texttt{components/pipeline.py} & Shared pipeline for CLI, UI, and tests \\
\bottomrule
\end{tabular}
\end{table}

\section{Technology Stack}
The project uses Python and standard-library tools only.

\begin{table}[H]
\centering
\caption{Technology Stack}
\begin{tabular}{ll}
\toprule
Tool & Purpose \\
\midrule
Python 3 & Core implementation language \\
\texttt{dataclasses} & AST and runtime helper structures \\
\texttt{tkinter} & Desktop graphical interface \\
\texttt{unittest} & Automated testing \\
\bottomrule
\end{tabular}
\end{table}

\section{Lexical Analyser}
The lexer reads the source character by character and emits tokens for
identifiers, literals, keywords, operators, and delimiters. It preserves line
and column information in each token to support precise diagnostics. Invalid
characters and unterminated strings are reported immediately with source
positions.

\section{Parser}
The parser is implemented as a handwritten recursive-descent parser. It builds
AST nodes for assignments, print statements, returns, blocks, conditional
statements, while-loops, function definitions, and function calls. Operator
precedence is encoded structurally through separate parsing functions for
factors, terms, expressions, and boolean expressions.

\section{Abstract Syntax Tree}
The AST is the main intermediate representation shared by the checker and the
interpreter. It contains nodes for literals, identifiers, binary operations,
assignments, block statements, \texttt{if}, \texttt{while}, function
definitions, function calls, \texttt{return}, and the built-in
\texttt{print()} function. Each node stores line and column metadata.

\section{Type Checker}
The type checker traverses the AST before execution. It uses the symbol table
to record variable and function information, infer types on first assignment,
validate arithmetic and comparison expressions, enforce Boolean conditions for
control flow, infer function parameter types from the first call, and infer
function return types from \texttt{return} statements. Programs that fail these
rules are rejected before interpretation.

\subsection{Implemented Type System Scope}
The implemented type system work covers the full semantic scope assigned for the
type system component. In presentation terms, this stage is responsible for
turning a syntactically valid AST into a semantically valid program that can be
executed with confidence:

\begin{itemize}
	\item static typing rules for arithmetic, comparison, assignment, control flow, and functions,
	\item type checking for expressions and statements,
	\item variable type inference from first assignment,
	\item function parameter inference from the first call,
	\item function return type inference from \texttt{return} statements,
	\item source-positioned type errors when a program violates a typing rule.
\end{itemize}

\section{Interpreter}
The interpreter is a tree-walking evaluator that executes the checked AST. It
uses nested runtime environments for variables and function calls. Assignments
store values in the current or nearest enclosing scope, \texttt{if} executes
only the selected branch, and \texttt{while} re-evaluates its condition until it
becomes false. Function calls create child environments, bind arguments by
value, and return through an internal return signal mechanism.

The built-in \texttt{print()} function outputs both value and runtime type, such
as \texttt{3 : Integer} or \texttt{"plc" : String}.

\subsection{Implemented Interpreter Scope}
The interpreter work covers all required runtime behaviours. In practical terms,
this stage is what gives the language observable behaviour from the user's point
of view:

\begin{itemize}
	\item assignment execution,
	\item \texttt{if} execution,
	\item \texttt{while} execution,
	\item function execution with value parameter passing,
	\item \texttt{print()} execution with immediate output,
	\item runtime integration through the shared pipeline used by the CLI, UI, and tests.
\end{itemize}

\section{Graphical User Interface}
The project includes a desktop UI built with \texttt{tkinter}. The interface
matches the requested layout: script input appears on the left, while output is
shown on the right. The right-hand side contains tabs for execution output,
tokens, AST, and inferred types. This allows the user to inspect each stage of
the language pipeline from one screen.
